---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
#### by Dan Jeffrey  

###Setup: 
```{r setup, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(lattice)
```
  
  
## Loading and preprocessing the data
```{r processing}
unzip("activity.zip")
rawData <- read.csv("activity.csv", colClasses = c("integer","Date","integer"))
str(rawData)
```
  
  
## What is mean total number of steps taken per day?
```{r hist1}
stepsByDate <- summarize(group_by(rawData, date), steps = sum(steps, na.rm=TRUE))
histogram(stepsByDate$steps, xlab="Number of Steps")
```
```{r stepsPerDay1, results="hide"}
meanStepsPerDay <- mean(stepsByDate$steps, na.rm = TRUE)
medianStepsPerDay <- median(stepsByDate$steps, na.rm = TRUE)
```
The mean number of steps per day is `r prettyNum(meanStepsPerDay, big.mark = ",")`.  
The median steps per day is `r prettyNum(medianStepsPerDay, big.mark = ",")`.  
  
  
## What is the average daily activity pattern?
```{r stepsByIntervalFigure }
stepsByInterval <- summarize(group_by(rawData, interval), 
                            avg=mean(steps, na.rm=TRUE))
with(stepsByInterval, plot(avg~interval, type="l", ylab="Average Number of Steps", xlab="Interval"))
maximumAverageSteps <- stepsByInterval[which.max(as.matrix(stepsByInterval)[,2]),]
```
The maximum number of steps per interval, on average, for all days was `r prettyNum(maximumAverageSteps[1,2], big.mark = ",")` in interval `r prettyNum(maximumAverageSteps[1,1], big.mark = ",")`.  

## Imputing missing values
```{r countMissing, results='hide'}
stepsNA <- is.na(rawData[,1])
nas <- sum(stepsNA)
```
The total number of rows with NA's is `r nas`.  Note that the Date and Interval fields contained no NA values.

The NA's are replaced with the mean values for the same interval to create a new dataframe with no NA values:
```{r fillMissing}
rawWithMean <- merge(rawData, stepsByInterval, by="interval")
filledDataset <- mutate(rawWithMean, steps = ifelse(is.na(steps), avg, steps))
```
  
#### With a "filled" data set: 
```{r hist2}
stepsByDate <- summarize(group_by(filledDataset, date), steps = sum(steps, na.rm=TRUE))
histogram(stepsByDate$steps, xlab="Number of Steps")
```
  
The above histogram shows that filling the missing values changes the data set significantly. It creates a spread of steps per day that is closer to a normal distribution.  
  
```{r getSTepsPerDay2, results="hide"}
meanStepsPerDay <- mean(stepsByDate$steps, na.rm = TRUE)
medianStepsPerDay <- median(stepsByDate$steps, na.rm = TRUE)
```
The mean number of steps per day is `r prettyNum(meanStepsPerDay, big.mark = ",")`.   
The median steps per day is `r prettyNum(medianStepsPerDay, big.mark = ",")`.  

## Are there differences in activity patterns between weekdays and weekends?
Assign a factor to each row that identifies weekends and weekdays:
```{r prepFilledWithWeekends}
filledWithWeekends <- mutate(filledDataset, day=factor(ifelse(weekdays(rawData$date) == "Saturday" | weekdays(rawData$date) == "Sunday", "weekend","weekday")))
str(filledWithWeekends)
```

```{r plotWeekends}
stepsByInterval <- summarize(group_by(filledWithWeekends, day, interval), 
                            avg=mean(steps))
xyplot(avg~interval | factor(day), data=filledWithWeekends, pch=19,
       lab="Wt/1,000",  ylab="Number of Steps",layout=c(1,2), xlab="Interval",
       type="l")
```

#### The End

